Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-02-24 ion_pairing_paper.tex
update table

Contrary to a small electrolyte that exhibit only dissociated ions, polyelectrolytes in solution are in general not fully dissociated, with a substantial fraction of the counterions bound to the polymer\cite{Manning1979}. The resulting net charge of the chain governs the physical properties of the polyelectrolytes solution\cite{Rubinstein1996}, first of all the ability of the polymer to dissolve. counterion condensation is the physical bounding or adsorption of counterions near the polymer chain. Factors influencing this process have been known since the 1880's but are still being rationalized. \citet{Hofmeister1888} was the first to propose a systematic ranking of ions, based on their propensity to promote the coalescence of egg white. This so-called Hofmeister series has since proved quite universally valid\cite{Zhang2010a}, including multi-charged polymers such as naturally occurring proteins or synthetic polyelectrolytes as well as charged colloidal particles\cite{Schwierz2010,Oncsik2015} or soft matter interfaces in general\cite{Leontidis2014}. However, depending on the nature of the polymer (hydrophilic or hydrophobic, anionic or cationic), the Hofmeister series can be direct (in short, well hydrated ions promote condensation) or reverse\cite{Schwierz2010,Oncsik2015}.

We have recently reported on a highly reversible poly(cation) based hydrogel\cite{Srour2014}, which formation relies on an new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate-terminated ATRP initiator, eventually affording an anion-terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.

In the present report, we investigate the role of the ion pair dissociation equilibria rate of the poly(cationic) body on the mechanical properties of the hydrogel. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (aromatic or not) of the cationic repeating unit and the associated counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}) varied along the Hofmeister series. In the second part and third part, we show respectively the qualitative change in gel formability and the quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a fourth part, we develop a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a polymer that is connected to three or more others) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}.%Finally, we discuss our results and possible applications of such smart hydrogels.

@@

Differential Scanning Calorimetry (DSC) thermograms  of all synthesized polymers also reveals marked differences in their properties, that can only be explained by differences in their couterion features. Thus, in the case polyelectrolytes with fluoride and chloride counterion, we observe no exo or endothermic transition below \SI{200}{\celsius}. By contrast, we observe a broad endothermic peak when iodide and bromide are used as counterions. Quite remarkably, we found a similar peak temperature (84-\SI{85}{\celsius}) for \ce{PIm+I-} and \ce{PPyr+I-}, see Figure~\ref{fig:dsc}. When bromide counterions are present, we observe a significant increase of the peak temperature (114-\SI{115}{\celsius}) but again, with a similar value between \ce{PIm+Br-} and \ce{PPyr+Br-} as shown in Figure~\ref{fig:dsc}. The overall shape and position of the DSC peaks is very reminiscent of previously reported data for various naturally occuring or synthetic poly(electrolytes)\cite{Li2005,Sarmento2006,Ostrowska-Czubenko2009a,Moin2015}. It is generally attributed to desorption of weakly bound water from the polymer network. %Its presence in the polymer samples where bromide and iodide are used as couterions suggests that adsorbed water could play a role in solvating the dissociated ion pairs in the latter. Conversely, its absence for chloride and fluoride counterions suggests that dissociation is not effective because of stronger ion pair interaction.

This difference in DSC profile can be understood by considering the difference in solvation of the different polymers in this study, as a function of the nature of the associated anion. A small anion like \ce{F-} can polarize a water molecule that is directly involved in hydrogen bonding with the ester functions displayed by our polymers\cite{Zhang2010a}. This water binding ability will decrease along the direct Hofmeister series $\ce{F-}>\ce{Cl-}>\ce{Br-}>\ce{I-}$. We are able to observe the end of this trend in the shift of peak position between bromide and iodide counterions, whereas in the case of chloride and fluoride, binding is too strong to observe water desorption. This hypothesis is further confirmed below, when looking at the properties of different polymers in aqueous dispersion.

@@

The above dependence of solubility on the nature of the counterions follows the direct Hofmeister series. The strong condensation of \ce{F-} on the chain decreases the effective positive charge of the polymer and thus reduces solubility. Conversely, larger, poorly hydrated ions like \ce{I-} are less adsorbed, resulting in more dissociated charges and higher solubility. As a consequence, while dissolution is observed for the four polymers with bromide and iodide counterions, in which ion dissociation takes place to a sufficient extent, the two polymers with fluoride counterions, for which couterion condensation is expected to be strong, cannot be dissolved.

However Hofmeister series alone cannot explain the difference between soluble \ce{PPyr+Cl-} and insoluble \ce{PIm+Cl-}. We have to suppose that counterions are more strongly condensed near imidazolium than near pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties. Because of their aromatic ring, imidazolium are particularly prone to promote a variety of supramolecular interactions which strongly contribute to their physicochemical properties, such as \ce{\pi+-\pi} or anion$-\pi^+$.  More recently, \ce{\pi+-\pi+} interactions have also been identified as a distinctive driving force for imidazolium dimerization\cite{Geronimo2011}. Because of strong intercoulombic repulsion, \ce{\pi+-\pi+} are primarily weaker than their \ce{\pi-\pi} counterparts. However it has been established that, in the presence of couterion, the stabilizing effect of \ce{\pi+-\pi+} interaction could reach magnitudes largely exceeding that of the latter\cite{Geronimo2011}. In particular, dimerization of imidazolium-chloride ion pairs have been the object of recent studies, and it has been shown that the involvement of the negatively charged counterion was essential in maximizing the stabilization of the \ce{\pi+-\pi+} complex by minimizing coulombic repulsion between the imidazolium moieties\cite{Matthews2014,Gao2015a}. It is likewise well documented that covalent incorporation of interacting group within a polymer chain strongly favours intramolecular interactions between these groups, by increasing their so-called effective molarity\cite{Li2003,Mulder2004,Huerta2013}. It is therefore not surprising that both effects participate in making the imidazolium-anion interaction stronger than pyrrolidinium at the detriment of solubility.

@@

On the other hand, if every chain has a single foreign head attached to its body, every polymer is linked to two others in a single file\bibnote{We neglect the possibility of the two charges of the phosphonate head to attach to two separate chains}. In addition, if we suppose that in order to minimize the inter-chain repulsion between charged cationic groups, heads are preferentially attached to the tail of their neighbour, we obtain a linear chain of effective polymerisation index $n N_0$ where $n$ is the number of polymers in the supramolecular queue. This behaviour bears a striking resemblance to pine processionary caterpillar behaviour first described by \citet{Fabre1916}: 
``They proceed in single file, in a continuous row, each touching with its head the rear of the one in front of it. [...] No Greek theoria winding its way to the Eleusinian festivals was ever more orderly. Hence the name of Processionary given to the gnawer of the pine.''

We thus have two limiting cases: (i) every chain has at least two heads attached and we have roughly $N_0$ monomers between cross-link points; (ii) every chain has at the most a single head attached and we have isolated supramolecular chains in the system. The former case should be observed when a significant number of ion-pairs along the polymer chain are dissociated providing many sites for a phosphonate head to bound to. The latter case should be observed at negligible charge dissociation. In between these two limiting cases, we should observe cross-link points separated by processions of $n N_0$ monomers as sketched in Figure~\ref{fig:dissociation}a.

As we will discuss in the following, all rheological features of the hydrogels studied here can  be rationalised on the basis of this ``processionary'' model. We base our analysis on two main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, and (ii) the critical strain $\gamma_c$ corresponding to the maximum of $G"$, that we consider as representative of the end of the linear regime and of the first broken cross-links\cite{Hyun2011}. By using this two parameters, we show that it is possible, with reasonable assumptions, to estimate microscopic parameters as the average number of chains between cross links $n$, the head-to-body bonding energy $E_c$ or the number of monomers between dissociated ion pairs $A$.

@@

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 790 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This larger cross-link ratio is consistent with a higher probability of attaching two or more heads on a highly dissociated body.

\begin{table}
\begin{tabular}{l|SS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$} & {$\sqrt{n}$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	1.	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.05&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	1.1	&	1.\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	45	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	7.	&	6.7\\
\ce{PIm+I-}	& 78	&	0.16	&	113	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	9.6	&	10.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	790	&	64	&	160.	&	3.4\cellcolor{gray!25}	&	28\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates inconsistent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

Finally we cross-check our model by linking the ion pair dissociation and the amount of physical cross-links by an independent argument. If we assume that the probability for an ion pair to dissociate is independent of the number of dissociated pairs already present on the polymer, we can describe the probability of having $k$ dissociated pairs on a polymer by a Poisson distribution of mean $\lambda = N_0/A$. Having 2 charges dissociated on a chain implies a cross-link. Therefore, the cross-link probability $1/n$ is proportional to $\lambda^2$, yielding $n \sim A^2$.

As shown in Figure~\ref{fig:dissociation}b and the last two columns of Table~\ref{tab:results} this relation describes very well our data. We can extrapolate this law for the case of \ce{PIm+Br-} and obtain $A = 28$. Inverting (Eq.~\ref{eq:A}) yields $R_0/R - 1 = 860\%$, indeed higher than the observed $\gamma_c\approx 804\%$. This confirms that in the \ce{PIm+Br-} strain sweep $\gamma_c$ actually results from the strength of the ionic bonds.

@@

Well hydrated anions and aromatic cation favour counterion condensation, resulting in too few charges to allow dissolution and gel formation. The hardest gels with the narrowest linear domain are obtained with aliphatic cationic moiety and poorly hydrated counterion at the other end of the Hofmeister series. This gives rise to maximal charge dissociation in water, facilitates charge pairing of the peripheral, iterative cations with the terminal anion and thus affords a very high density of cross-links. In between, in the case of aliphatic cations with a counterion in the middle of the Hofmeister series (\ce{Cl-}) or in the case of aromatic cation and poorly hydrated anion, charge dissociation is limited, polymers associate in processions and cross-links are few. We thus obtain soft, highly deformable gels.

>>>>

Contrary to a small electrolyte for which full ion pairs dissociation is generally achieved in water, polyelectrolytes in solution are in general not fully dissociated, with a substantial fraction of the counterions bound to the polymer\cite{Manning1979}. The resulting net charge of the chain governs the physical properties of the polyelectrolyte solution\cite{Rubinstein1996}, first of all the ability of the polymer to dissolve. Counterion condensation is the physical bounding or adsorption of counterions near the polymer chain. Factors influencing this process have been known since the 1880's but are still not fully understood. \citet{Hofmeister1888} was the first to propose a systematic ranking of ions, based on their propensity to promote the coalescence of egg white. This so-called Hofmeister series has since proved quite universally valid\cite{Zhang2010a}, including multi-charged polymers such as naturally occurring proteins or synthetic polyelectrolytes as well as charged colloidal particles\cite{Schwierz2010,Oncsik2015} or soft matter interfaces in general\cite{Leontidis2014}. However, depending on the nature of the polymer (hydrophilic or hydrophobic, anionic or cationic), the Hofmeister series can be direct (in short, well hydrated ions promote condensation) or reverse\cite{Schwierz2010,Oncsik2015}.

We have recently reported on a highly reversible poly(cation) based hydrogel\cite{Srour2014}, which formation relies on a new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate-terminated ATRP initiator, eventually affording an anion-terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.

In the present report, we investigate the role of the couterion condensation on the mechanical properties of such hydrogels. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (aromatic or not) of the cationic repeating unit and the associated halide counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}) varied along the Hofmeister series. In the second and third parts, we show respectively the qualitative change in gel formability and the quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a fourth part, we rationalize this behaviour by proposing a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a point where three or more polymers meet) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}.%Finally, we discuss our results and possible applications of such smart hydrogels.

@@

Differential Scanning Calorimetry (DSC) thermograms  of all synthesized polymers also reveals marked differences in their properties, that can only be explained by differences in their couterion features. Thus, in the case polyelectrolytes with fluoride and chloride counterion, we observe no exo or endothermic transition below \SI{200}{\celsius} (featureless thermograms, not shown). By contrast, we observe a broad endothermic peak when iodide and bromide are used as counterions, see Figure~\ref{fig:dsc}. Quite remarkably, we found a similar peak temperature (84-\SI{85}{\celsius}) for \ce{PIm+I-} and \ce{PPyr+I-}. When bromide counterions are present, we observe a significant increase of the peak temperature (114-\SI{115}{\celsius}) but again, with a similar value between \ce{PIm+Br-} and \ce{PPyr+Br-} as shown in Figure~\ref{fig:dsc}. The overall shape and position of the DSC peaks is very reminiscent of previously reported data for various naturally occuring or synthetic poly(electrolytes)\cite{Li2005,Sarmento2006,Ostrowska-Czubenko2009a,Moin2015}. It is generally attributed to desorption of weakly bound water from the polymer network. %Its presence in the polymer samples where bromide and iodide are used as couterions suggests that adsorbed water could play a role in solvating the dissociated ion pairs in the latter. Conversely, its absence for chloride and fluoride counterions suggests that dissociation is not effective because of stronger ion pair interaction.

This difference in DSC profile can be understood by considering the water-binding ability of the different polymers in this study. Since the peak temperature does not depend on the presence of imidazolium or pyrrolidinium, water-binding is probably due to the only part of the size groups able to accept hydrogen bonds from water: the ester function. Following \citet{Zhang2010a}, we propose that these bonds can be enhanced if the involved water molecule is polarized by solvating an anion. The polarization, and thus the water-binding ability will decrease along the direct Hofmeister series $\ce{F-}>\ce{Cl-}>\ce{Br-}>\ce{I-}$. We are able to observe the end of this trend in the shift of peak position between bromide and iodide counterions, whereas in the case of chloride and fluoride, binding is too strong to observe water desorption. This hypothesis is further confirmed below, when looking at the properties of different polymers in aqueous dispersion.

@@

The above dependence of solubility on the nature of the counterions follows the direct Hofmeister series. The strong condensation of \ce{F-} on the chain decreases the effective positive charge of the polymer and thus reduces solubility. Conversely, bulkier and softer ions like \ce{I-} are less adsorbed, resulting in more dissociated charges and higher solubility. As a consequence, while dissolution is observed for the four polymers with bromide and iodide counterions, in which ion dissociation takes place to a sufficient extent, the two polymers with fluoride counterions, for which couterion condensation is expected to be strong, cannot be dissolved.

However Hofmeister series alone cannot explain the difference between soluble \ce{PPyr+Cl-} and insoluble \ce{PIm+Cl-}. This constitutes a strong indication that counterions are more strongly condensed near imidazolium than near pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties. Because of their aromatic ring, imidazolium ions are particularly prone to promote a variety of supramolecular interactions which strongly contribute to their physicochemical properties, such as \ce{\pi+-\pi} or anion$-\pi^+$.  More recently, \ce{\pi+-\pi+} interactions have also been identified as a distinctive driving force for imidazolium dimerization\cite{Geronimo2011}. Because of strong intercoulombic repulsion, \ce{\pi+-\pi+} interactions are primarily weaker than their \ce{\pi-\pi} counterparts. However it has been established that, in the presence of couterions, the stabilizing effect of \ce{\pi+-\pi+} interaction could reach magnitudes largely exceeding that of the latter\cite{Geronimo2011}. In particular, dimerization of imidazolium-chloride ion pairs have been the object of recent studies, and it has been shown that the involvement of the negatively charged counterion was essential in maximizing the stabilization of the \ce{\pi+-\pi+} complex by minimizing coulombic repulsion between the imidazolium moieties\cite{Matthews2014,Gao2015a}. It is likewise well documented that covalent incorporation of interacting groups within a polymer chain strongly favours intramolecular interactions between these groups, by increasing their so-called effective molarity\cite{Li2003,Mulder2004,Huerta2013}. It is therefore not surprising that both effects participate in making the imidazolium-anion interaction stronger than pyrrolidinium at the detriment of solubility.

@@

On the other hand, if every chain has a single foreign head attached to its body, every polymer is linked to two others in a single file\bibnote{We neglect the possibility of the two charges of the phosphonate head to attach to two separate chains}. In addition, if we suppose that in order to minimize the inter-chain repulsion between charged cationic groups, heads are preferentially attached to the tail of their neighbour, we obtain a linear chain of effective polymerisation index $n N_0$ where $n$ is the number of polymers in the supramolecular queue. This behaviour is somewhat evocative of that of the pine processionary caterpillar memorably described by \citet{Fabre1916}: 
``They proceed in single file, in a continuous row, each touching with its head the rear of the one in front of it. [...] No Greek theoria winding its way to the Eleusinian festivals was ever more orderly. Hence the name of Processionary given to the gnawer of the pine.''

We thus have two limiting cases: (i) every chain has at least two heads attached and we have roughly $N_0$ monomers between cross-link points; (ii) every chain has at the most a single head attached and we have isolated supramolecular chains in the system. The former case should be observed when a significant number of ion-pairs along the polymer chain are dissociated providing a significant probability for multiple phosphonate heads binding on a single polymer chain. Conversely, the latter case should be observed at low charge dissociation. In between these two limiting cases, we should observe cross-link points separated by processions of $n N_0$ monomers as sketched in Figure~\ref{fig:dissociation}a.

As we will discuss in the following, all rheological features of the hydrogels studied here can  be rationalised on the basis of this ``processionary'' model. We base our analysis on two main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, and (ii) the critical strain $\gamma_c$ corresponding to the maximum of $G"$, that we consider as representative of the end of the linear regime and of the first broken cross-links\cite{Hyun2011}. By using these two parameters, we show that it is possible, with reasonable assumptions, to estimate microscopic parameters as the average number of chains between cross-links $n$, the head-to-body bonding energy $E_c$ or the number of monomers between dissociated ion pairs $A$.

@@

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 880 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This larger cross-link ratio is consistent with a higher probability of attaching two or more heads on a highly dissociated body.

\begin{table}
\begin{tabular}{l|SS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$} & {$\sqrt{n}$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.08	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	10.6	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.15&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	11.7	&	1.\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	51	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	76.2	&	6.7\\
\ce{PIm+I-}	& 78	&	0.16	&	123	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	101.8	&	10.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	882	&	64	&	160.	&	36.5\cellcolor{gray!25}	&	28\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates inconsistent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

Finally we cross-check our model by linking the ion pair dissociation and the amount of physical cross-links by an independent argument. A second charge dissociated on a chain creates a possible binding site for a second head-to-body association and thus a cross-link. Therefore the probability to have a cross link is proportional to the probability of having a second charge dissociation. If we assume that the probability for an ion pair to dissociate is independent of the number of dissociated pairs already present on the polymer, we can describe the probability of having $k$ dissociated pairs on a polymer by a Poisson distribution of mean $\lambda = N_0/A$. Therefore, the cross-link probability $1/n$ is proportional to $\lambda^2$, yielding $n \sim A^2$.

As shown in Figure~\ref{fig:dissociation}b and the last two columns of Table~\ref{tab:results} this relation describes very well our data with a prefactor close to 1. We can extrapolate this law for the case of \ce{PIm+Br-} and obtain $A = 28$. Inverting (Eq.~\ref{eq:A}) yields $R_0/R - 1 = 860\%$, indeed higher than the observed $\gamma_c\approx 804\%$. This confirms that in the \ce{PIm+Br-} strain sweep $\gamma_c$ actually results from the strength of the ionic bonds.

@@

Strongly interacting anions (i.e. small and hard halides, like fluoride) and aromatic cations favour counterion condensation, resulting in too few charges to allow dissolution and gel formation. The hardest gels with the narrowest linear domain are obtained with aliphatic cationic moiety and counterion at the other end of the Hofmeister series that are less bound to the polymer. This gives rise to maximal charge dissociation in water, facilitates charge pairing of the peripheral, iterative cations with the terminal anion and thus affords a very high density of cross-links. In between, in the case of aliphatic cations with a counterion in the middle of the Hofmeister series (\ce{Cl-}) or in the case of aromatic cations and poorly interacting anion, charge dissociation is limited, polymers associate in processions and cross-links are few. We thus obtain soft, highly deformable gels.
